Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-08-23 pot/LaTeX/jsm2014/jsm2014.tex
New plots, tweaks to simstudy

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture.
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels.
      \item e.g. Probability of exceeding a certain threshold level
    \end{itemize}
    \item Spatial methods borrow information across space to estimate spatial correlation and make predictions by Kriging at unknown locations.
    \item Want to explore similar methods for extremes.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Standard analysis - Peaks over threshold}
  \begin{itemize}  \setlength{\itemsep}{0.5em}
      \item Incorporates more data than block maxima
      \item Select a threshold, $T$, and use the Generalized Pareto distribution (GPD) to model the exceedances.
      \item Temporal dependence may be an issue between observations (e.g. flood levels don't dissipate overnight)
  \end{itemize}
\end{frame}

@@

\begin{frame}{Gaussian spatial model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item In geostatistics $Y(\bs)$ are often modeled using a Gaussian process with mean function $\mu(\bs)$ and covariance function
$\rho(\bh)$.
    \item Model properties:
    \begin{itemize}
      \item Nice computing properties (closed-form likelihood)
    \item For a Gaussian spatial model $\lim_{c \rightarrow \infty} \chi(\bh) = 0$ regardless of the strength of the correlation in the bulk of the distribution.
    \item Tail is not flexible (Gaussian is light tailed)
    \end{itemize}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item \alert{Conditioned} on $z_t$ and $\sigma^2_t$, $Y_t(\bs)$ is a Gaussian spatial model
    \item Can use standard geostatistical methods to fit this model.
    \item Predictions can be made through Kriging.
    \item \alert{Marginalizing} over $z_t$ and $\sigma^2_t$ (via MCMC),
    \begin{align*}
      Y_t(\bs) \sim \text{skew-t}(\mu, \Sigma^*, \alpha, \text{df}=2a)
    \end{align*}
    where
    \begin{itemize}
    	\item $\mu$ is the location
	\item $a$, $b$ are the IG parameters for $\sigma^2_t$
	\item $\Sigma^* = \frac{ b }{ a } \Sigma$ is a scale matrix, and $\Sigma$ is a \Matern covariance matrix
	\item $\alpha \in \calR$ controls the skewness
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
      \item Model properties
    \begin{itemize}
    	\item Has flexible tail controlled by skewness $\alpha$ and degrees of freedom $2a$
    	\item For a skew-$t$ distribution $\lim_{c \rightarrow \infty} \chi(\bh) > 0$ (Padoan, 2011)
    	\item Computation that is on the order of Gaussian computation
    \end{itemize}
    \item For this distribution, $\chi(\bh)$ shows asymptotic dependence that does not approach 0 as $\bh \rightarrow \infty$
   \item This occurs because all observations (near and far) share the same $z_t$ and $\sigma_t^2$.
    \item We deal with this through a daily random partition (similar to Huser and Davison).
  \end{itemize}
\end{frame}

\begin{frame}{Daily random partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Daily random partition allows $z_t$ and $\sigma^2_t$ to vary by site.
    \begin{align*}
      Y_t(\bs) = X_t(\bs) \beta + \alpha z_t(\bs) + \sigma(\bs) v_t(\bs)
    \end{align*}
    \item Consider a set of daily knots $\bw_{tk} \sim$ Uniform that define a random daily partition
    $P_{t1}, \ldots, P_{tK}$ such that
    \begin{align*}
      P_{tk} = \{s : k = \argmin_\ell|| \bs - \bw_{t\ell}|| \}
    \end{align*}
    \item For $\bs \in P_{tk}$
    \begin{align*}
      z_t(\bs) &= z_{tk}\\
      \sigma^2_t(\bs) &= \sigma^2_{tk}
    \end{align*}
    \item Within each partition $Y_t(\bs)$ has the same MV skew-t distribution as before.
  \end{itemize}
\end{frame}

@@

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps:
    \begin{enumerate}[1.]
      \item Impute censored data below $T$
      \item Update parameters with standard random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Data analysis}
  \centering
  \begin{figure}
    \includegraphics[width=1\linewidth]{./plots/ozone-day.pdf}
    \caption{Max 8-hour ozone measurements at 85 sites in NC, SC, and GA for days 5 and 34.}
   \end{figure}

@@

\begin{frame}{Five-fold cross-validation results}
  \begin{table}[htbp]
    \small
    \centering
    \begin{tabular}{l|c|r|rrrrr}
          \multicolumn{3}{c}{\ } & \multicolumn{5}{c}{$\tau$}\\
           \hline
  Marginal & $K$ & $T$  & 0.950 & 0.980 & 0.990 & 0.995 & 0.999\\
  \hline
Gaussian 	& 1 & 0 	& 39.820 & 17.539 & 9.167 & 4.720 & 1.057\\
$t$		& 1 & 0 	& {\bf 31.008} & {\bf 13.898} & 7.229 & {\bf 3.405} & 0.879\\
$t$		& 3 & 0	& 31.213 & 13.920 & {\bf 7.218} & 3.498 & 0.918\\
$t$		& 1 & 0.9	& 32.221 & 14.519 & 7.549 & 3.604 & 0.896\\
$t$		& 3 & 0.9 	& 38.842 & 16.781 & 8.434 & 4.180 & 1.020\\
skew-$t$	& 1 & 0	& 31.845 & 14.542 & 7.533 & 3.645 & {\bf 0.844}\\
skew-$t$	& 1 & 0.9	& 32.132 & 14.296 & 7.484 & 3.497 & 0.890\\
skew-$t$	& 3 & 0	& 33.653 & 15.453 & 8.119 & 4.338 & 1.188\\
skew-$t$	& 3 & 0.9 	& 32.157 & 14.727 & 7.794 & 3.825 & 0.917\\
\hline
    \end{tabular}
    \caption{Brier score for predicting exceedance of $c = \hat{q}(\tau)$ from five-fold cross-validation ($\times 1000$)}
  \end{table} \vspace{-1.5em}
  \begin{itemize}
  	\item Quantile score results are similar.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Simulation study}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 6 different data settings:
    \begin{itemize}
    	\item Gaussian vs $t$ vs skew-$t$ marginal distribution
        \item $K=1$ partition vs $K=5$ partitions
    \end{itemize}
    \item Preliminary results are inconclusive.
  \end{itemize}
\end{frame}

>>>>

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels
      \item e.g. Probability of exceeding a certain threshold level
    \end{itemize}
    \item Spatial methods borrow information across space to estimate spatial correlation and make predictions by Kriging at unknown locations
    \item Want to explore similar methods for extremes
  \end{itemize}
\end{frame}

@@

\begin{frame}{Standard analysis - Peaks over threshold}
  \begin{itemize}  \setlength{\itemsep}{0.5em}
      \item Incorporates more data than block maxima
      \item Select a threshold, $T$, and use the Generalized Pareto distribution (GPD) to model the exceedances
      \item Temporal dependence may be an issue between observations (e.g. flood levels don't dissipate overnight)
  \end{itemize}
\end{frame}

@@

\begin{frame}{Gaussian spatial model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item In geostatistics $Y(\bs)$ are often modeled using a Gaussian process with mean function $\mu(\bs)$ and covariance function
$\rho(\bh)$.
    \item Model properties:
    \begin{itemize}
      \item Nice computing properties (closed-form likelihood)
    \item For a Gaussian spatial model $\lim_{c \rightarrow \infty} \chi(\bh) = 0$ regardless of the strength of the correlation in the bulk of the distribution
    \item Tail is not flexible (Gaussian is light tailed)
    \end{itemize}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item \alert{Conditioned} on $z_t$ and $\sigma^2_t$, $Y_t(\bs)$ is a Gaussian spatial model
    \item Can use standard geostatistical methods to fit this model
    \item Predictions can be made through Kriging
    \item \alert{Marginalizing} over $z_t$ and $\sigma^2_t$ (via MCMC),
    \begin{align*}
      Y_t(\bs) \sim \text{skew-t}(\mu, \Sigma^*, \alpha, \text{df}=2a)
    \end{align*}
    where
    \begin{itemize}
    	\item $\mu$ is the location
	\item $a$, $b$ are the IG parameters for $\sigma^2_t$
	\item $\Sigma^* = \frac{ b }{ a } \Sigma$ is a scale matrix, and $\Sigma$ is a \Matern covariance matrix
	\item $\alpha \in \calR$ controls the skewness
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
      \item Model properties
    \begin{itemize}
    	\item Has flexible tail controlled by skewness $\alpha$ and degrees of freedom $2a$
    	\item For a skew-$t$ distribution $\lim_{c \rightarrow \infty} \chi(\bh) > 0$ (Padoan, 2011)
    	\item Computation that is on the order of Gaussian computation
    \end{itemize}
    \item For this distribution, $\chi(\bh)$ shows asymptotic dependence that does not approach 0 as $\bh \rightarrow \infty$
   \item This occurs because all observations (near and far) share the same $z_t$ and $\sigma_t^2$
    \item We deal with this through a daily random partition (similar to Huser and Davison)
  \end{itemize}
\end{frame}

\begin{frame}{Daily random partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Daily random partition allows $z_t$ and $\sigma^2_t$ to vary by site
    \begin{align*}
      Y_t(\bs) = X_t(\bs) \beta + \alpha z_t(\bs) + \sigma(\bs) v_t(\bs)
    \end{align*}
    \item Consider a set of daily knots $\bw_{tk} \sim$ Uniform that define a random daily partition
    $P_{t1}, \ldots, P_{tK}$ such that
    \begin{align*}
      P_{tk} = \{s : k = \argmin_\ell|| \bs - \bw_{t\ell}|| \}
    \end{align*}
    \item For $\bs \in P_{tk}$
    \begin{align*}
      z_t(\bs) &= z_{tk}\\
      \sigma^2_t(\bs) &= \sigma^2_{tk}
    \end{align*}
    \item Within each partition $Y_t(\bs)$ has the same MV skew-t distribution as before
  \end{itemize}
\end{frame}

@@

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps:
    \begin{enumerate}[1.]
      \item Impute censored data below $T$
      \item Update parameters with standard random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible
  \end{itemize}
\end{frame}

@@

\begin{frame}{Data analysis}
  \centering
  \begin{figure}
    \includegraphics[width=1\linewidth]{./plots/ozone-day.pdf}
    \caption{Max 8-hour ozone measurements at 85 sites in NC, SC, and GA for days 5 and 34}
   \end{figure}

@@

\begin{frame}{Five-fold cross-validation results}
  \begin{table}[htbp]
    \small
    \centering
    \begin{tabular}{l|c|r|rrrrr}
          \multicolumn{3}{c}{\ } & \multicolumn{5}{c}{$\tau$}\\
           \hline
  Marginal & $K$ & $T$  & 0.950 & 0.980 & 0.990 & 0.995 & 0.999\\
  \hline
Gaussian 	& 1 & 0 	& 39.820 & 17.539 & 9.167 & 4.720 & 1.057\\
$t$		& 1 & 0 	& {\bf 31.008} & {\bf 13.898} & 7.229 & {\bf 3.405} & 0.879\\
$t$		& 3 & 0	& 31.213 & 13.920 & {\bf 7.218} & 3.498 & 0.918\\
$t$		& 1 & 0.9	& 32.221 & 14.519 & 7.549 & 3.604 & 0.896\\
$t$		& 3 & 0.9 	& 38.842 & 16.781 & 8.434 & 4.180 & 1.020\\
skew-$t$	& 1 & 0	& 31.845 & 14.542 & 7.533 & 3.645 & {\bf 0.844}\\
skew-$t$	& 1 & 0.9	& 32.132 & 14.296 & 7.484 & 3.497 & 0.890\\
skew-$t$	& 3 & 0	& 33.653 & 15.453 & 8.119 & 4.338 & 1.188\\
skew-$t$	& 3 & 0.9 	& 32.157 & 14.727 & 7.794 & 3.825 & 0.917\\
\hline
    \end{tabular}
    \caption{Brier score for predicting exceedance of $c = \hat{q}(\tau)$ from five-fold cross-validation ($\times 1000$)}
  \end{table} \vspace{-1.5em}
  \begin{itemize}
  	\item Quantile score results are similar
  \end{itemize}
\end{frame}

@@

\begin{frame}{Simulation study}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 6 different data settings:
    \begin{itemize}
    	\item Gaussian vs $t$ vs skew-$t$ marginal distribution
        \item $K=1$ partition vs $K=5$ partitions
    \end{itemize}
    \item Preliminary results are inconclusive
  \end{itemize}
\end{frame}
